Learning Libraries of Subroutines for Neurally--Guided Bayesian Program Learning

### 2018-05-18 nips_2018.tex
Josh introduction

Imagine you are asked  to edit some text, and
told that you should change the text ``Nancy FreeHafer'' to
``Dr. Nancy''.  From this example, you likely infer that ``Jane
Goodall'' goes to ``Dr. Jane'', drawing upon your prior knowledge of
text, like that words are separated by spaces. Few-shot learning problems
like these
are commonplace in both human and machine learning.
We cast these few-shot learning problems as each being a program synthesis problem.
The programming languages and AI communities have built many successful program synthesis algorithms,
spanning text editing (e.g., FlashFill~\cite{gulwani2011automating}), motor programs~\cite{lake2015human}, procedural graphics~\cite{ellis2017learning}, planning
procedures~\cite{devlin2017neural}, and many others.  However, the
success of these systems hinges upon a carefully
hand-engineered \textbf{Domain Specific Language (DSL)}.  DSLs
impart prior knowledge of a domain
by providing a restricted set
of finely-tuned programming primitives: for
%that are finely tuned to the problems in the domain: for
text editing,
these are primitives like appending and splitting on characters.
In this work,
we consider the problem of
building agents that solve program learning tasks,
and also the problem of acquiring
the prior knowledge necessary to quickly
solve these tasks (Figure~\ref{firstPageFigure}).
Our solution is an algorithm that learns a DSL
while jointly training a neural network to help write programs in the learned DSL.
\begin{figure}[h]
  \begin{minipage}[c]{0.68\textwidth}
\hspace{-0.5cm}      \begin{tabular}{r|c|}\cmidrule{2-2}
  \sc{Task}&
  \begin{tabular}{rcl}
  Nancy FreeHafer&$\longrightarrow$& Dr. Nancy\\
% Andrew Cencici&$\longrightarrow$&Dr. Andrew\\
 Jane Goodall&$\longrightarrow$&???
  \end{tabular}\\\cmidrule{2-2}
  \sc{Program}&$f(\text{\code{s}})\,=\,$\code{(}$f_0$\code{ "Dr." (}$f_1$\code{ s ' '))}\\\cmidrule{2-2}
    \sc{DSL}
&\begin{tabular}{l}
    $f_0($\code{a}$,$\code{b}$) \,=\, $\code{(foldr a b ($\lambda$ (x y) (cons x y)))}\\
    \hspace{0.5cm}($f_0$: \emph{Appends lists (of characters))}\\
    $f_1($\code{s}$,$\code{c}$) \,=\, $\code{(foldr s s ($\lambda$ (x a)}\\
    \phantom{$f_1($\code{s}$,$\code{c}$) \,=\, $\code{  }}\code{(if (= c x) nil (cons x a))))}\\
        \hspace{0.5cm}($f_1$: \emph{Take characters from }\code{s}\emph{ until }\code{c}\emph{ reached)}
    \end{tabular}\\\cmidrule{2-2}
  \end{tabular}
  \end{minipage}%
  \begin{minipage}[c]{0.31\textwidth}
      \caption{\textbf{Task}: Few-shot learning problem. Model solves tasks by writing \textbf{program}s, and jointly learns a library of reusable subroutines that are shared across multiple tasks, called a \textbf{Domain Specific Language (DSL)}. Program writing is guided by a neural network trained jointly with the library.}\label{firstPageFigure}
    \end{minipage}
\end{figure}

%% We think of
%% solutions to these tasks as being well represented by programs,
%% and so our problem can be stated as follows:
%% how should an agent learn to write programs?
We take inspiration from two sources:
(1) Good software engineers compose libraries
of reusable subroutines
that are shared across
related programming tasks.
Returning to Figure~\ref{firstPageFigure},
a good text editing library should support appending strings and splitting on spaces --
exactly the prior knowledge needed to solve the task in Figure~\ref{firstPageFigure}.
(2) Skilled human programmers
can quickly recognize what kinds of programming idioms and library routines would be useful for solving
the task at hand, even if they cannot instantly work out the details.
We weave these ideas together into
an algorithm called \system (\textbf{Co}mpress/\textbf{Co}mpile/\textbf{Sea}rch),
which takes as input a collection of
programming \textbf{tasks},
and then jointly
solves three problems:
(1) Searching for programs that solve the tasks;
(2) Composing a library (DSL) of
domain-specific
subroutines -- which allow the agent to
more compactly write programs in the domain,
and (3) Training a neural network
to recognize
which DSL components are
useful for which kinds of tasks.
Together, the DSL and neural net
encode the domain specific knowledge needed to
quickly write programs.

%% synthesize programs for task domains such as string transformations,
%% list processing, and robot navigation and planning. However, all these
%% approaches -- symbolic, neural and neural-symbolic -- rely upon a
%% hand-engineered \emph{Domain-Specific Language} (DSL). DSLs contain an
%% inventory of restricted programming primitives, encoding domain-specific
%% knowledge about the space of programs. In
%% practice
%% we often have only a few input/output examples for each
%% program to be induced, and thus success often hinges on having a good
%% DSL that provides a crucial inductive bias for what would otherwise be an
%% unconstrained search through the space of all computable functions.
%% Here we ask, to what extent can we dispense with such highly
%% hand-engineered domain-specific languages?

%% We propose \emph{learning} the DSL by inducing a library of domain--specific
%% subroutines. We consider the setting where we have a collection of
%% related programming tasks, each specified by a set of input/output
%% examples. %% We do not assume that the tasks are annotated with
%% %% ground-truth programs.
%% Starting from a weaker or more
%% general library of primitives, we give an algorithm for constructing a richer, more powerful,
%% and better-tuned DSL.

%% Our algorithm is called \textbf{Explore/Compress/Compile} (\systemEnding),
%% because it
\system iterates through three different steps: a
\textbf{Search} step uses the DSL and neural network to explore the space of programs,
searching for ones that solve the tasks; a \textbf{Compress} step
modifies the structure of the DSL by discovering regularities across programs found during search; and a \textbf{Compile} step, which
improves the search procedure by training a neural network to
write programs in the current DSL, in the spirit of ``amortized'' or
``compiled'' inference~\cite{le2016inference}.
We call the neural net a \textbf{recognition model} (c.f. Hinton 1995~\cite{hinton1995wake}).
The learned DSL
distills commonalities across programs that solve tasks, helping
the agent solve related programming problems. The neural
recognition model ensures that searching for programs remains tractable
even as the DSL (and hence the search space for programs) expands.
%% Because any model may be encoded as a (deterministic or probabilistic) program,
%% we carefully delineate the scope of problems considered here.
We think of \system as learning to solve the kinds of problems that humans
can solve relatively quickly -- once they acquire the relevant domain expertise.
These correspond to short programs -- if you have an expressive DSL.
Even with a good DSL, program search may be intractable, so
we amortize the cost of search by training a neural network to
assist the search procedure.

We apply \system to three domains:
list processing; text editing (in the style of FlashFill~\cite{gulwani2011automating}); and symbolic regression.
 For each of these we initially provide a generic
 set of programming primitives.
Our algorithm then constructs
its own DSL for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).
\begin{table}[hb]
  \makebox[\textwidth][c]{
    \scriptsize
  \tabcolsep=4pt
  \renewcommand\code\texttt
  \renewcommand\codechar[1]{\texttt{"#1"}}
  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{>{\hspace{-1em}}c<{\hspace{-1em}}>{\hspace{-2.5em}}c<{\hspace{-0.8em}}>{\hspace{-1em}}c<{\hspace{-1em}}}
    \toprule
    {\normalsize List Functions}&{\normalsize Text Editing}&{\normalsize Symbolic Regression}\\\midrule
    {\tabcolsep=4pt
      \begin{tabular}{cc}
        \begin{tabular}{c}
          \code{[7\, 0\, 2]}$\to $\code{[7\, 0\, 2\, 9]}         \\
          \code{[9\, 3]}$\to $\code{[9\, 3\, 9]} \\
          \code{$f(\ell) = $}\code{($f_1$ $\ell$ 9)}             \\
          ~                                                  \\
          ~                                                  \\
          \code{[2\, 7\, 8\, 1]}$\to $\code{8}               \\
          \hspace{0.15cm}\code{[3\, 19\, 14]}$\to $\code{19}                \\
          \code{$f(\ell) = $}\code{($f_2$ $\ell$)}
        \end{tabular}
        &
        \hspace{-0.3cm}\begin{tabular}{c}
          \code{[7\, 3]}$\to $\code{False}                              \\
          \hspace{0.3cm}\code{[3]}$\to $\code{False}                    \\
          \hspace{-0.3cm}\code{[9\, 0\, 0]}$\to $\code{True\phantom{e}} \\
          \hspace{0.3cm}\code{[0]}$\to $\code{True\phantom{e}}                        \\
          \hspace{-0.3cm}\code{[0\, 7\, 3]}$\to $\code{True\phantom{e}}                \\
          \code{$f(\ell) = $}\code{($f_3$ $\ell$ 0)}
        \end{tabular}
      \end{tabular}
    }
    &
    \hspace{-0.5cm}\begin{tabular}{c}
      +106 769-438$\to $106.769.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
      +83 973-831$\to $83.973.831\\
      $f(\text{\code{s}}) = $\code{(}$f_0$\code{  \codechar{.} \codechar{,} }\\
      \hspace{1.25cm}\code{($f_0$ \codechar{.} \codechar{ }}\\
      \hspace{1.4cm}\code{(cdr s)))}\\
      ~\\
      Temple Anna H $\to $TAH\\
      Lara Gregori$\to $LG\\
      $f(\text{\code{s}}) = $\code{(}$f_2$\code{ s)}\\
    \end{tabular}
    &
    \begin{tabular}{cc}
      \includegraphics[width = 3em]{figures/functions/4.png}&
      \includegraphics[width = 3em]{figures/functions/146}\\
      \code{$f($x$) = $($f_1$ x)}&    \code{$f($x$) = $($f_6$ x)}\\
      ~\\
      \includegraphics[width = 3em]{figures/functions/112.png}&
        \includegraphics[width = 3em]{figures/functions/92.png}
      \\
      \code{$f($x$) = $($f_4$ x)}&    \code{$f($x$) = $($f_3$ x)}\\

@@

@@

 \section{The \system Algorithm}

 Our goal is to induce a DSL while finding programs solving each of the
tasks.  We take inspiration primarily from the
Exploration-Compression algorithm for bootstrap
learning~\cite{Dechter:2013:BLV:2540128.2540316}.  Exploration-Compression
alternates between exploring the space of solutions to a set of tasks,
and compressing those solutions to suggest new search primitives
for the next exploration stage.
We extend these ideas into an inference strategy that iterates
through three steps: a \textbf{Search} step uses the current DSL and
recognition model to search for programs that solve the tasks.  The
\textbf{Compress} and \textbf{Compile} steps update the DSL and the
recognition model, respectively.  Crucially, these steps bootstrap 
each other:
\begin{comment}
  \begin{wrapfigure}{r}{0.5\textwidth}
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs};
    \node[align = center](DSL) at (3,1) {DSL};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model};

    \draw [->,thick] (synthesis.-120) to[out = -150,in = 60] node[below,rotate = 45,align = center]{{\footnotesize Trains}\\{\footnotesize (Compression)}} (DSL.30);
    \draw [->,thick] (synthesis.-60) to[out = -30,in = 120] node[below,rotate=-45,align = center]{{\footnotesize Trains}\\{\footnotesize (Compilation)}} (recognitionModel.150);
    \draw [->,thick] (DSL.east) to[out = -30,in = 210] node[above, align = center]{{\footnotesize Trains}\\{\footnotesize (Compilation)}} (recognitionModel.west);

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Inductive bias}\\\footnotesize{(Exploration)}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white,align = center]{{\footnotesize Makes tractable}\\{\footnotesize (Exploration)}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
  \caption{\system solves for programs, the DSL, and a neural network (recognition model). Each of these steps iteratively bootstraps off of the others.}  \label{feeding}\vspace{-0.4cm}
\end{wrapfigure}
  \end{comment}
\\\noindent \textbf{Search: Solving tasks.}  Our program search
is informed by both the DSL and the recognition model. When these
improve, we can solve more tasks.
\\\noindent\textbf{Compression: Improving the DSL.} We induce the DSL
from the programs found in the search phase, aiming
to maximally compress (or, raise the prior probability of) these
programs.  As we solve more tasks, we hone in on DSLs that
more closely match the domain.
\\\noindent \textbf{Compilation: Learning a neural recognition model.}  We
update the recognition model by training on two data sources: samples
from the DSL (as in the Helmholtz Machine's ``sleep'' phase), and
programs found by the search procedure during search. As the DSL
improves and as search finds more programs, the recognition model gets
more data to train on, and better data.

\begin{comment}
Sec.~\ref{mathematicalFraming} frames this 3-step procedure as probabilistic inference.
Sec.~\ref{explorationSection} explains how we search for programs that solve the tasks;
Sec.~\ref{recognitionSection} explains how we train a neural network to  search for programs; and
Sec.~\ref{grammarInductionSection} explains how we induce a DSL from programs.
  \end{comment}
\subsection{Hierarchical Bayesian Framing}\label{mathematicalFraming}

@@

\begin{figure}\centering
  \includegraphics[width = 3.5cm]{figures/list.eps}
  \includegraphics[width =  3.5cm]{figures/textLearningCurve.eps}        
  \includegraphics[width = 3.5cm]{figures/rationalCurve.eps} 
  \caption{Learning curves for \system both with (blue) and without (red) the recognition model. Solid lines: \% holdout testing tasks solved. Dashed lines: Average solve time.}\label{learningCurves} 
\end{figure}

>>>>

Much of everyday human thinking and learning can be understood in
terms of program induction: constructing a procedure that maps inputs
to desired outputs, based on observing example input-output pairs.
People can induce programs flexibly across many different domains, and
remarkably, often from just one or a few examples.  For instance, if
shown that a text-editing program should map ``Jane Morris Goodall''
to ``J. M. Goodall'', we can guess it maps ``Richard Erskine Leakey''
to ``R. E. Leakey''; if instead the first input mapped to
``Dr. Jane'', ``Goodall, Jane'', or ''Morris'', we might have guessed
the latter should map to ``Dr. Richard'', ``Leakey, Richard'', or
``Erskine'', respectively.

The FlashFill system ~\cite{gulwani2011automating} developed by
Microsoft researchers and now embedded in Excel solves problems such
as these and is probably the best known practical program-induction
algorithm, but researchers in programming languages and AI have built
successful program induction algorithms for many applications, 
such as handwriting recognition and generation ~\cite{lake2015human},
procedural graphics ~\cite{ellis2017learning}, question
answering~\cite{johnson2017clevr} and robot motion
planning~\cite{devlin2017neural}, to name just a few.
These systems work in different ways, but most hinge upon having a
carefully engineered \textbf{Domain Specific Language (DSL)}.  This is
especially true for systems such as FlashFill that aim to induce a
wide range of programs very quickly, in a few seconds or less.  DSLs
constrain the search over programs with strong prior knowledge in the
form of a restricted set of programming primitives tuned to the needs
of the domain: for text editing, these are operations like appending
strings and splitting on characters.

In this work, we consider the problem of building agents that learn to
solve program induction tasks, and also the problem of acquiring the
prior knowledge necessary to quickly solve these tasks in a new domain.
Representative problems in three domains are shown in 
Table~\ref{initialExampleDSL}.  Our solution is an algorithm that
grows or boostraps a DSL while jointly training a neural network to help write
programs in the increasingly rich DSL.  
%
% (JBT: MAYBE INSERT FIGURE FROM ICML PAPER?)
%

\begin{table}[t!]
  \makebox[\textwidth][c]{
    \scriptsize
  \tabcolsep=4pt
  \renewcommand\code\texttt
  \renewcommand\codechar[1]{\texttt{"#1"}}
  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{>{\hspace{-1em}}c<{\hspace{-1em}}>{\hspace{-2.5em}}c<{\hspace{-0.8em}}>{\hspace{-1em}}c<{\hspace{-1em}}}
    \toprule
    {\normalsize List Functions}&{\normalsize Text Editing}&{\normalsize Symbolic Regression}\\\midrule
    {\tabcolsep=4pt
      \begin{tabular}{cc}
        \begin{tabular}{c}
          \code{[7\, 0\, 2]}$\to $\code{[7\, 0\, 2\, 9]}         \\
          \code{[9\, 3]}$\to $\code{[9\, 3\, 9]} \\
          \code{$f(\ell) = $}\code{($f_1$ $\ell$ 9)}             \\
          ~                                                  \\
          ~                                                  \\
          \code{[2\, 7\, 8\, 1]}$\to $\code{8}               \\
          \hspace{0.15cm}\code{[3\, 19\, 14]}$\to $\code{19}                \\
          \code{$f(\ell) = $}\code{($f_2$ $\ell$)}
        \end{tabular}
        &
        \hspace{-0.3cm}\begin{tabular}{c}
          \code{[7\, 3]}$\to $\code{False}                              \\
          \hspace{0.3cm}\code{[3]}$\to $\code{False}                    \\
          \hspace{-0.3cm}\code{[9\, 0\, 0]}$\to $\code{True\phantom{e}} \\
          \hspace{0.3cm}\code{[0]}$\to $\code{True\phantom{e}}                        \\
          \hspace{-0.3cm}\code{[0\, 7\, 3]}$\to $\code{True\phantom{e}}                \\
          \code{$f(\ell) = $}\code{($f_3$ $\ell$ 0)}
        \end{tabular}
      \end{tabular}
    }
    &
    \hspace{-0.5cm}\begin{tabular}{c}
      +106 769-438$\to $106.769.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
      +83 973-831$\to $83.973.831\\
      $f(\text{\code{s}}) = $\code{(}$f_0$\code{  \codechar{.} \codechar{,} }\\
      \hspace{1.25cm}\code{($f_0$ \codechar{.} \codechar{ }}\\
      \hspace{1.4cm}\code{(cdr s)))}\\
      ~\\
      Temple Anna H $\to $TAH\\
      Lara Gregori$\to $LG\\
      $f(\text{\code{s}}) = $\code{(}$f_2$\code{ s)}\\
    \end{tabular}
    &
    \begin{tabular}{cc}
      \includegraphics[width = 3em]{figures/functions/4.png}&
      \includegraphics[width = 3em]{figures/functions/146}\\
      \code{$f($x$) = $($f_1$ x)}&    \code{$f($x$) = $($f_6$ x)}\\
      ~\\
      \includegraphics[width = 3em]{figures/functions/112.png}&
        \includegraphics[width = 3em]{figures/functions/92.png}
      \\
      \code{$f($x$) = $($f_4$ x)}&    \code{$f($x$) = $($f_3$ x)}\\

@@

Because any computable learning problem can in principle be cast as
program induction, it is important to delimit our focus.  In contrast
to computer assisted programming~\cite{solar2008program} or genetic
programming~\cite{DBLP:books/daglib/0070933}, our goal is not to automate software
engineering, to learn to synthesize large bodies of code, or to learn
complex programs starting from scratch.  Ours is a basic AI goal:
capturing the human ability to learn to think flexibly and efficiently
in new domains -- to learn what you need to know about a domain so you
don't have to solve new problems starting from scratch.  We are
focused on the kinds of problems that humans can solve relatively
quickly, once they acquire the relevant domain expertise.  These
correspond to tasks solved by short programs -- if you have an
expressive DSL.  Even with a good DSL, program search may be
intractable;
%
% , and adding new library routines to the DSL only broadens the
% search space;
%
so we amortize the cost of search by training a
neural network to assist the search procedure.

Our algorithm takes inspiration from several ways that skilled human
programmers have learned to code: Skilled coders build libraries of
reusable subroutines that are shared across related programming tasks,
and can be composed to generate increasingly complex and powerful
subroutines.  In text editing, a good library should support routines
for splitting on characters, but also specialize these routines to
split on particular characters such as spaces or commas that are
frequently used to delimit substrings across tasks.  Skilled coders
also learn to recognize what kinds of programming idioms and library
routines would be useful for solving the task at hand, even if they
cannot instantly work out the details.  In text editing, one might
learn that if outputs are consistently shorter than inputs, removing
characters is likely to be part of the solution; if every output
contains a constant substring (e.g., ``Dr.''), inserting or appending
that constant string is likely to be a subroutine.

Our \system (\textbf{(S)}search/\textbf{(C})ompress/\textbf{(C})ompile)
algorithm incorporates these insights by iterating through three steps. 
The \textbf{Search} step takes a given set of \textbf{tasks}, typically
several hundred, and searches for compact programs that solve these tasks
guided by the current DSL and neural network.
The \textbf{Compress} step grows the library (or DSL) of
domain-specific subroutines which allow the agent to more compactly
write programs in the domain; it modifies the structure of the DSL by
discovering regularities across programs found during search, compressing
them to distill out common code fragments across successful programs.
The \textbf{Compile} step improves the search procedure by training a neural network to
write programs in the current DSL, in the spirit of ``amortized'' or
``compiled'' inference~\cite{le2016inference}.%%  (CITE Goodman and Stuhlmuller,
%% and others too on amortized inference.)

The learned DSL effectively encodes a prior on programs likely to
solve tasks in the domain, while the neural net looks at the example
input-output pairs for a specific task and produces a ``posterior''
for programs likely to solve that specific task.  The neural network
thus functions as a \textbf{recognition model} supporting a form of
approximate Bayesian program induction, jointly trained with a
\textbf{generative model} for programs encoded in the DSL, in the
spirit of the Helmholtz machine ~\cite{hinton1995wake}). The
recognition model ensures that searching for programs remains
tractable even as the DSL (and hence the search space for programs)
expands.

We apply \system to three domains:
list processing; text editing (in the style of FlashFill~\cite{gulwani2011automating}); and symbolic regression.
 For each of these we initially provide a generic
 set of programming primitives.
Our algorithm then constructs
its own DSL for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).

@@

 \section{The \system Algorithm}
 We first mathematically describe our 3-step algorithm as
 an inference procedure for a hierarchical Bayesian model (Section~\ref{mathematicalFraming}),
  and then describe each step algorithmically in detail (Section~\ref{explorationSection}-\ref{grammarInductionSection}).

 \subsection{Hierarchical Bayesian Framing}\label{mathematicalFraming}

@@

\begin{figure}\centering
  \includegraphics[width = 4.5cm]{figures/list.eps}
  \includegraphics[width = 4.5cm]{figures/textLearningCurve.eps}        
  \includegraphics[width = 4.5cm]{figures/rationalCurve.eps}   
  \caption{Learning curves for \system both with (blue) and without (red) the recognition model. Solid lines: \% holdout testing tasks solved. Dashed lines: Average solve time.}\label{learningCurves} 
\end{figure}
